\documentclass[a4paper, 12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb,amsfonts,bbm,bm}
\usepackage[parfill]{parskip}
\usepackage{hyperref}

\title{trend-calculus}
\author{Johannes Graner \and Raazesh Sainudiin \and whomsoeverCoauthor(s)}

\date{August 2020}

\begin{document}

\maketitle

\section{Introduction}

A trend in a financial time series can be thought of as "a perceived tendency of financial markets to move in a particular direction over time".\footnote{\url{https://en.wikipedia.org/wiki/Market_trend}}
In other words, if the trend is up, the price or value tends to increase and if the trend is down, the price or value tends to decrease. 

However, most time series fluctuate quite rapidly so some smoothing has to happen to detect what is a trend and what is just random noise.

%What does trend mean in time series... what are trend reversals and why do we want to detect trend-reversals? % keep it brief and point to wikipedia or other publicly available resources using \footnote{\url{...}}


\section{Data}

A typical dataset for this type of application is financial data, either historical data or some stream of trades happening right now. Such a time series typically contains at least three things: 

\begin{itemize}
    \item A ticker, some label for what is being traded.
    \item A timestamp that tells us when the data is from. This can be the timestamps for individual trades or a timestamp for when some aggregation has finished. An example of such an aggregation is the average over the last minute.
    \item A value, usually the closing price or average closing price for aggregated data.
\end{itemize}

Such data is available from e.g. \url{http://www.histdata.com/}.\footnote{\url{https://github.com/lamastex/FX-1-Minute-Data}}

%What is a typical dataset here and what specific ones are we using -- yfinance, fx1M, etc. Also point o github lamastex forks of these 

\section{Algorithm}

Trend Calculus\footnote{\url{https://github.com/ByteSumoLtd/TrendCalculus-lua/tree/master/benchmark}} is invented by Andrew Morgan and is implemented is Apache Spark and made compatible with Spark Structured Streaming\footnote{\url{https://github.com/lamastex/spark-trend-calculus}} by Johannes Graner under supervision of Raazesh Sainudiin.

GitHub link to spark-trend-calculus... and Andrew's Lua version...

\section{Mathematical interpretation}

\subsection{Algorithm}

The input to the algorithm is a time series with $N$ observations, 

\[ T = \{ p_i = (t_i, x_i) \in \mathbb{R}^2 : t_i < t_{i+1} \forall i = 1, \dots, N-1 \} \]

and a window size $s \in [2, N] \cap \mathbb{Z}$.

To create the windows in the algorithm, we define the function $f_s : T^s \to T^2$,

\begin{align*}
    f_s(p_{i_1}, \dots, p_{i_s}) = ( & \min_t (\min_x \{ (t_{i_j}, x_{i_j}) : j = 1,\dots,s \}), \\
    & \max_t (\max_x \{ (t_{i_j}, x_{i_j}) : j = 1,\dots,s \}) )
\end{align*}

which finds the earliest minimum and latest maximum among the points $p_{i_1},\dots,p_{i_s}$.

Assume that $N$ is divisible by $s$ (otherwise, only consider $N - \lceil\frac{N}{s} \rceil$ points of the series). Let $n = \frac{N}{s} \in \mathbb{Z}_{\ge 2}$ and let 

\begin{equation*}
    W = \{ w_k = (l_k, h_k) = f_s(p_{1+sk},\dots,p_{s+sk}) : k = 1,\dots,n\}. 
\end{equation*}

Then $W$ is the set of windows that summarize the time series $T$. For each $w_k \in W$, define the trend for that window as 

\begin{equation*}
    \tau_k = \begin{cases}
        1, & k = 1 \\
        \tau(w_{k-1}, w_k), & k > 1
    \end{cases}
\end{equation*}

where $\tau : W^2 \to \{-1,0,1\}$ is given by

\begin{equation*}
    \tau(w_{k-1}, w_k) = sign(sign \circ \pi_x(l_k - l_{k-1}) + sign \circ \pi_x(h_k - h_{k-1}))
\end{equation*}

$\pi_x : \mathbb{R}^2 \to \mathbb{R}$ is the projection $(t,x) \mapsto x$.

To correct the trend when $\tau_k = 0$, let 

\begin{equation*}
    w_k^* = \begin{cases}
        w_{k - \#\{j : j<k, t_j = 0\}}, & \tau_k \ne 0 \\
        (\min_x\{a,b\}, \max_x\{a,b\}), & \tau_k = 0
    \end{cases}
\end{equation*}

where $a = \max_t\{l_{k-1}, h_{k-1}\}, b = \min_t\{l_k, h_k\}$. This creates intermediate windows at every point where the trend is 0 and inserts them at the correct places in the sequence $(w_k^*)$. The length of this sequence is $n^* = n + \#\{k : \tau_k = 0\}$. This sequence is then used to define the corrected trends $\tau_k^*$.

\begin{equation*}
    \tau_k^* = \begin{cases}
        1, & k = 1 \\
        \tau(w_{k-1}^*, w_k^*), & k>1 \; and \;  \tau(w_{k-1}^*, w_k^*) \ne 0 \\
        \tau_{k-1}^*, & otherwise
    \end{cases}
\end{equation*}

This ensures that every trend is non-zero. If a zero trend is found with the corrected windows as well, the last non-zero trend is carried forward. The non-zero trends are used to define reversal of trends $r_k$ as $r_1 = 1$, $r_k = sign(\tau_{k-1}^* - \tau_k^*)$ and let

\begin{equation*}
    T^*(k) = \begin{cases}
        \{ \min_t\{l_1, h_1\} \}, & k = 1 \\
        \{ l_{k-1} \}, & r_k = 1 \; and \; k > 1 \\
        \{ h_{k-1} \}, & r_k = -1 \; and \; k > 1 \\
        \varnothing, & r_k = 0 \; and \; k > 1
    \end{cases}
\end{equation*}

If $T^*(k) \ne \varnothing$, then $T^*(k)$ is a singleton and we call the element in $T^*(k)$ a \textit{reversal point}.

This can be used to define a new time series $T^{(1)} = \cup_{k=1}^{n^*} T^*(k) \subset T$ which contain only the points of T where a trend reversal happens. The reversal points can be used to determine the \textit{trend} at every point since the reversal points are the points at which the trend changes. The trend at a point $p_i$ is $trend_i = (-1)^{\#\{p_j \in T^{(1)} : t_j \le t_i\} + 1}$, i.e. the trend at $p_1$ is 1 and switches sign at the reversal points.

\subsection{Multiple iterations}

The output of the algorithm, $T^{(1)}$, can then be used for another iteration to get $T^{(2)} \subset T^{(1)}$. This can be repeated until $\#T^{(m)} < s$, at which point we get $T^{(m^*)} = \varnothing \; \forall m^* > m$ and a decreasing sequence of sets $\mathcal{T}$ can be defined as $\mathcal{T} = (T, T^{(1)}, \dots, T^{(m)})$ where $T \supset T^{(1)} \supset, \dots, T^{(m)}$. We therefore define the \textit{order} of a reversal point $p$ to be $\rho^*(p) = max (\{ k : p \in T^{(k)} \}$. A point $p$ that is not a reversal point, i.e. $p \in T \backslash T^{(1)}$, then it has reversal order 0, $\rho^*(p) = 0$.

Since $\mathcal{T}$ is a decreasing sequence of sets, one can augment the original time series $T$ with information about every reversal that has occurred thus far. Each point in the augmented time series $T_{tc}$ is $(t_i,x_i,\rho_i) \in \mathbb{R}^2 \times \{ -m,\dots,m \}$ where $\rho_i = trend_i \cdot \rho^*(p_i)$.

So, if a point is a reversal of order $r$, then it will be augmented with $r$ if it is a local minimum and $-r$ if it is a local maximum. If a point is not a reversal, it will be augmented with $0$.

We then see that $T^{(k)} = \{ p : \rho^*(p) \ge k \} $ so $T_{tc}$ contains all information necessary to recreate $\mathcal{T}$ as well as information about the type of reversal. It therefore also contains all information necessary to determine the trend at every point of $T$.

\section{Modeling}



\section{State of Structured Streams}

The real problem...

\section{Showcases}

A github.io page with showcases for Trend Calculus is available at \url{https://lamastex.github.io/spark-trend-calculus-examples/}.

\end{document}
